---
title: 'Data Science: Probability'
output:
  word_document: default
  pdf_document: default
  html_notebook: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 3) # report 3 significant digits
```

```{r library , message=FALSE, warning=FALSE, include=TRUE}
library(tidyverse)
```


## Questions 1 and 2: ACT scores, part 1
The ACT is a standardized college admissions test used in the United States. The four multi-part questions in this assessment all involve simulating some ACT test scores and answering probability questions about them.

For the three year period 2016-2018, ACT standardized test scores were approximately normally distributed with a mean of 20.9 and standard deviation of 5.7. (Real ACT scores are integers between 1 and 36, but we will ignore this detail and use continuous values instead.)

---

First we'll simulate an ACT test score dataset and answer some questions about it.

Set the seed to 16, then use **rnorm** to generate a normal distribution of 10000 tests with a mean of 20.9 and standard deviation of 5.7. Save these values as **act_scores**. You'll be using this dataset throughout these four multi-part questions.

In order to generate a normal distribution of $10^{4}$ tests with mean of 20.9 and standard deviation of 5.7, we need to use the function **rnorm(n,avg,sd)**. To do this first we to setup the **set.seed**.

```{r set-seed}
x <- 16
set.seed(x, sample.kind = "Rounding")
```
```{r act-scores}
size_sample <- 10^4
avg <- 20.9
sd <- 5.7
act_scores <- rnorm(size_sample, avg, sd)
```

Let's plot the **act_score** to check it up is looks like a normal distribution.

```{r plot-act_scores}
data.frame(act_scores = act_scores) %>%
  ggplot(aes(x = act_scores)) +
  geom_histogram(color = "black", fill = "blue", binwidth = 2) +
  geom_vline(xintercept = avg, color = "red", linetype = "dashed",    size = 1.5)
```

* Question 1a
What is the mean of act_scores ?

```{r mean-act_scores}
mean(act_scores)
```

* Question 1b

What is the standard deviation of act_scores?

```{r sd-act_scores}
sd(act_scores)
```

* Question 1c

A perfect score is 36 or greater (the maximum reported score is 36).


In act_scores, how many perfect scores are there out of $10^{4}$ simulated tests? 
```{r perfect-scores}
perfect_scores <- 36
data.frame(act_scores = act_scores) %>%
  filter(act_scores >= perfect_scores) %>%
  summarize(count = n(), sum(act_scores, na.rm = TRUE))
```
* Question 1d

In act_scores, what is the probability of an ACT score greater than 30

Solution 1d:
**$1 - Pr(actscores > 30) = 1 - pnorm(30, avg, sd)$
```{r 1d-probability}
1 - pnorm(30, avg, sd)
```

* Question 1e

In act_scores, what is the probability of an ACT score less than or equal to 10?


Solution 1e:

The probability of ACT score $Pr(actscores <= 10) = pnorm(10, avg, sd)$

```{r Probability}
pnorm(10, avg, sd)
```
---

* Question 2

Set x equal to the sequence of integers 1 to 36. Use dnorm to determine the value of the probability density function over x given a mean and standard deviation of 5.7; save the result as f_x. Plot x against f_x
```{r dnorm_plot}

x <- seq(1:36)

data.frame(x, f_x = dnorm(x, 20.9, 5.7)) %>%
  ggplot(aes(x = x, y = f_x)) + geom_line(color = "blue") +
  geom_vline(xintercept = 20.9, color = "red", linetype = "dashed",    size = 1.5) 
```


---
## Question 3 and 4: ACT scores, part2

In this 3-part question, you will convert ran ACT scores to Z-scores and answer some questions about them.

* Question 3a

Convert act_scores to Z-scores.Recall from Data Visualization that to standardize values (convert values into Z-scores, that is, values distributed with a mean of 0 and standard deviation of 1), you must subtract the mean and then divide by the standard deviation. Use the mean and standard deviation of **act_scores**, not the original values to generate random test scores.

```{r z-scores}

z_scores <- (act_scores - mean(act_scores))/sd(act_scores)
mean(z_scores > 2)
```

* Question 3b

What ACT score value corresponds to 2 standard deviations above the mean (z = 2)?

```{r z-scores-3b}

# The score value can be calculated using the following code:
2*sd(act_scores) + mean(act_scores)
```

* Question 3c

A Z-score of 2 corresponds roughly to the 97.5th percentile

Use **qnorm** to determine the 97.5th percentile of normally distributed data with the mean and standard deviation observed in act_scores.

What is the 97.5th percentile of act_scores?
```{r qnorm-act_scores}

qnorm(0.975, avg, sd)
```

* Question 4a

What is the minimum integer score such that the probability of that score or lower is at least 0.95?

```{r 4a}

cdf <- sapply(1:36, function (x){
  mean(act_scores <= x)
})
min(which(cdf >= .95))
```

* Question 4b

Use qnorm to determine the expected 95th percentile, the value for the probability of receiving that score or lower is 0.95, given a mean score of 20.9 and standard deviation of 5.7

What is the 95th percentile of act_scores?
```{r 4b}

qnorm(0.95, 20.9, 5.7)
```

* Question 4c

Make a vector containing the quantiles for p <- seq(0.01, 0.99, 0.01), the 1s through 99th percentiles of the act_scores data. Save these as sample_quantiles.

```{r 4c}

p <- seq(0.01, 0.99, 0.01)
sample_quantiles <- quantile(act_scores, p)
names(sample_quantiles[max(which(sample_quantiles < 26))])
```

* Question 4d
Make a corresponding set of theoretical quantiles using qnorm over the interval p <- seq(0.01, 0.99, 0.01) with mean 20.9 and standard deviation 5.7. Save these as theoretical_quantiles. Make a QQ-plot graphing sample_quantiles on the y-axis versus theoretical_quantiles on the x-axis.

```{r 4d}
p <- seq(0.01, 0.99, 0.01)
sample_quantiles <- quantile(act_scores, p)
theoretical_quantiles <- qnorm(p, 20.9, 5.7)
qplot(theoretical_quantiles, sample_quantiles) + geom_abline() 
```

