---
title: "discrete-probability-assessment-part1"
output:
  word_document: default
  pdf_document: default
  html_notebook: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 3) # report 3 significant digits
```

## Assessment: Discrete Probability
### Question 1: Olympic running


Libraries that will be used for the calculations in this assessment. This assessment has 3 sections, all related to *Discrete Probability*.

```{r Libraries, message=FALSE, warning=FALSE}

library(gtools)
library(tidyverse)

```
  In the 200m dash finals in the Olympics, 8 runners compete for 3 medals 
(order matters). In the 2012 Olympics, 3 of the 8 runners were from Jamaica 
and the other 5 were from different countries. The three medals were all won 
by Jamaica (Usain Bolt, Yohan Blake, and Warren Weir).

Use the information above to help you answer the following four questions.

**Question 1a** - How many different ways can the 3 medals be distributed across 8 runners?

**Solution 1a :**
To answer this question we need to remember in this kind of competion the orders matters, for the first runner, he will get gold medal, and so on, it seems we 
have to use permutations function to answer this question.

```{r Answer 1a }

all_medals <- nrow(permutations(8,3)) # if orders matters, we must use permutations
                                      # nrow will give us the number of the observations
all_medals
```

---

**Question 1b** - How many different ways can the three medals be distributed among the 3 runners from Jamaica?

**Solution 1b :**
To answer this question we can use the same approach for previous question, the only diference, we have to make the permutations between the runners from Jamaica only.

```{r Answer 1b }

nrow(permutations(3,3)) # all the permutations between 3 runners from the same country

```

---

**Question 1c** - What is the probability that all 3 medals are won by Jamaica?


**Solution 1c:** 


S = all different ways the 3 medals distributed between 8 runners *(answer question 1a)


$n_{runners from Jamaica}$ = all ways the Jamaica runners can win


$Pr(runners from Jamaica)$ = **probability that all 3 medals are won by Jamaica**

```{r Answer 1c }
runners_Jamaica <- nrow(permutations(3,3)) # Jamaica cases
all_8_runners <- nrow(permutations(8,3))   # all cases

Pr_runners_Jamaica <- runners_Jamaica / all_8_runners
Pr_runners_Jamaica  # our answer

```
---

**Question 1d** - Run a Monte Carlo simulation on this vector representing of the 8 runners in this race:

```{r question 1d vector of runners}
runners <- c("Jamaica", "Jamaica", "Jamaica", "USA", "Ecuador", "Netherlands",
             "France", "South Africa")
```
Calculate the probability that all the runners are from Jamaica.


**Solution 1d :**

In order to create a Monte Carlo Simulation, we need to 3 steps:

1.Trials:


We need to create a trial which it will represent the sample from the runners vectors, then take 3 samples at time.

```{r trials}
runners_medals <- sample(runners, size = 3)
```
2.Success


Here we must get all winners from Jamaica.

```{r success}
all(runners_medals == "Jamaica")
```
3. Monte Carlo Simulation


Using the definitions we create on step 1 and step 2, here we code the simulation
```{r Monte Carlo Simulation}
set.seed(1)
B <- 10^4 # number of simulations 
medals <- replicate(B, {
  runners_medals <- sample(runners, size = 3)
  all(runners_medals %in% "Jamaica")
})
(Pr_runners_Jamaica <- mean(medals)) # show all the probabilities for the Jamaica runners
Pr_runners_Jamaica
```
