---
title: "discrete-probability-assessment-part3"
output:
  html_notebook: default
  word_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 3) # report 3 significant digits
```

## Questions 3 and 4: Esophageal cancer and alcohol/tobacco use, part1


Case-control studies help determine whether certain exposures are associated with outcomes such as developing cancer. The built-in dataset **esoph** contains data from a case-control study in France comparing people with esophageal cancer (cases, counted in **ncases**) to people without esophageal cancer (controls, counted in **ncontrols**) that are carefully matched on a variety of demographic and medical characteristics. The study compares alcohol intake in grams per day (**alcgp**) and tobacco intake in grams per day (**tobgp**) across cases and controls grouped by age range (**agegp**).


The dataset is available in base R and can be called with the variable name **esoph**:
```{r dataset}

head(esoph)
```

You will be using this dataset to answer the following four multi-part questions (Questions 3-6).


You may wish to use the **tydyverse** package:
```{r message=FALSE, warning=FALSE}
library(tidyverse)
```


---

The following three parts have you explore some basic characteristics of the dataset.


Each row contains a combination of age group, alcool consumption group, and tobacco consumption group, followed by the number of cancer cases and number of controls for study subjects in that combined grouping.

---

* Question3a:
How many groups are in the study?

**Solution3a:**
You can find the number of groups using
```{r 3a}

nrow(esoph)
```

* Question3b:
How many cases are there?*(tip: Save this value as all_cases for later problems)*
**Solution3b:**
```{r 3b}
all_cases <- sum(esoph$ncases)
all_cases
```

* Question3c:
How many controls are there?*(tip: Save this value as all_controls for later problems)*
**Solution3c:**
```{r 3c}

all_controls <- sum(esoph$ncontrols)
all_controls
```
---

The following four parts ask you to explore some probabilities within this dataset related to alcohol and tobacco consumption.


---


* Question4a:

What is the probability that a subject in the highest alcohol consuption group is a cancer case?


**Solution4a:**
You can find the probability using this code:
```{r 4a}

esoph %>%
  filter(alcgp == "120+") %>%
  summarize(ncases = sum(ncases), ncontrols = sum(ncontrols)) %>%
  mutate(p_case = ncases / (ncases + ncontrols)) %>%
  pull(p_case)
```

* Question4b:*

What is the probability that subject in the lowest consumption group is a cancer case?


**Solution4b:**
You can find the probability using this code:
```{r 4b}

esoph %>%
  filter(alcgp == "0-39g/day") %>%
  summarize(ncases = sum(ncases), ncontrols = sum(ncontrols)) %>%
  mutate(p_case = ncases / (ncases + ncontrols)) %>%
  pull(p_case)
```


* Question4c:

Given that a person is a case, what is the probability that they smoke 10g or more a day?


**Solution4c:**
You can find the probability using this code:
```{r 4c}

tob_cases <- esoph %>%
  filter(tobgp != "0-9g/day") %>%
  pull(ncases) %>%
  sum()

tob_cases/all_cases
```


* Question4d*:

Given that a person is a control, what is the probability that they smoke 10g or more a day?


**Solution4d:**
You can find the probability using this code:
```{r 4d}

tob_controls <- esoph %>%
  filter(tobgp != "0-9g/day") %>%
  pull(ncontrols) %>%
  sum()

tob_controls/all_controls
```
---


#Questions 5 and 6: Esophageal cancer and alcohol/tobacco use, part 2


The following four parts look at probabilities related to alcohol and tobacco consumption among the cases.


---

* Question 5a

For cases, what is the probability of being in the highest alcohol group?


**Solution5a:**
The probability can be calculated using the following code:
```{r 5a}

high_alc_cases <- esoph %>%
  filter(alcgp == "120+") %>%
  pull(ncases) %>%
  sum()

p_case_high_alc <- high_alc_cases/all_cases
p_case_high_alc
```

* Question 5b

For cases, what is the probability of being in the highest tobacco group?


**Solution5b:**
The probability can be calculated using the following code:
```{r 5b}

high_tob_cases <- esoph %>%
  filter(tobgp == "30+") %>%
  pull(ncases) %>%
  sum()

p_case_high_tob <- high_tob_cases/all_cases
p_case_high_tob
```

* Question5c:

For cases, what is the probability of being in the highest alcohol group **and** the highest tobacco group?


**Solution5c:**
The probability can be calculated using the following code:
```{r 5c}

high_alc_tob_cases <- esoph %>%
  filter(alcgp == "120+" & tobgp == "30+") %>%
  pull(ncases) %>%
  sum()

p_case_high_alc_tob <- high_alc_tob_cases/all_cases
p_case_high_alc_tob
```


* Question5d:

For cases, what is the probability of being in the highest alcohol group **or** the highest tobacco group?


**Solution5d:**
The probability can be calculated using the following code:
```{r 5d}
p_case_either_highest <- p_case_high_alc + p_case_high_tob - p_case_high_alc_tob
p_case_either_highest
```

---


The following six parts look at probabilities related to alcohol and tobacco consumption among the controls and also compare the cases and the controls.


---


* Question6a:

For controls, what is the probability of being in the highest alcohol group?


**Solution6a:**
The probability can be calculated using the following code:
```{r 6a}

high_alc_controls <- esoph %>%
  filter(alcgp == "120+") %>%
  pull(ncontrols) %>%
  sum()

p_control_high_alc <- high_alc_controls/all_controls
p_control_high_alc
```

* Question 6b:

How many time more likely are cases than controls to be in the highest alcohol group?


**Solution6b:**
The probability can be calculated using the following code:
```{r 6b}

p_case_high_alc/p_control_high_alc
```

* Question6c:

For controls, what is the probability of being in the highest tobacco group?

**Solution6c:**
The probability can be calculated using the following code:
```{r 6c}

high_tob_controls <- esoph %>%
  filter(tobgp == "30+") %>%
  pull(ncontrols) %>%
  sum()

p_control_high_tob <- high_tob_controls/all_controls
p_control_high_tob
```

* Question6d:

For controls, what is the probability of being in the highest alcool group **and** the highest tobacco group?

**Solution6d:**
The probability can be calculated using the following code:
```{r 6d}

high_alc_tob_controls <- esoph %>%
  filter(alcgp == "120+" & tobgp == "30+") %>%
  pull(ncontrols) %>%
  sum()

p_control_high_alc_tob <- high_alc_tob_controls/all_controls
p_control_high_alc_tob
```


* Question6e:

For controls, what is the probability of being in the highest alcohol group **or** the highest tobacco group?


**Solution6e:**
The probability can be calculated using the following code:
```{r 6e}

p_control_either_highest <- p_control_high_alc + p_control_high_tob - p_control_high_alc_tob
p_control_either_highest
```


* Question6f:

How many times more likely are cases than controls to be in the highest alcohol group **or** the highest tobacco group?


**Solution6f:**
This calculated using the following code:
```{r 6f}

p_case_either_highest/p_control_either_highest
```

